\vspace{-0.2in} Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-06-05 main.tex
Applied fixes from parents, and Ray

Among its many applications, $\ell_1$-min has been recently used to reformulate
image-based face recognition as a sparse representation problem
\cite{WrightJ2009-PAMI}.  If we stack the pixels of the training images of $K$ subject
classes into the columns of matrices $(A_1\in\Re^{d\times n_1}, \cdots, A_K\in\Re^{d\times n_K})$, combine
the matrices into a larger matrix $A = [A_1, \cdots, A_K]\in\Re^{d\times n}$, and arrange the pixels of a new
query image into a vector $\bb\in\Re^d$, \emph{sparsity-based
classification} (SBC) solves the following minimization problem:
\begin{equation}
\min_{\xx, \ee} \| \xx \|_1 + \|\ee\|_1 \quad \subj \quad \bb = A \xx + \ee.
\label{eq:l1min_denoise}
\end{equation}
If the sparsest solutions for $\x$ and $\e$ are recovered, $\ee$ provides a
means to compensate for pixels that are corrupted due to occlusion of some part of the query
image, and the dominant nonzero coefficients of $\xx$ reveal the membership of
$\bb$ based on the training image labels associated with $A$. 
%Since $A$ contains many images per user taken under different illuminations, 
%$A_i\x_i$ acts as a linear illumination model for the test image $\bb$.

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of subjects or more will likely require computation to be
distributed over a cluster or network of computers. These systems are not addressed here.
The source code of our implementations, which target NVIDIA Fermi GPUs
as well as Intel x86-64 CPUs, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.\vspace{-0.1in}}
\label{fig:pipeline}
\end{figure}

@@

Our discussion and experiments will address the most common hardware
configuration for engineering workstations: a motherboard with two quad-core
processors, and a PCI card with a single high-end GPU.  
Recognition involving more than a few hundred subjects with contemporary hardware
would require a server (or cluster) configuration with an expandable 
number of processors.  We will not address efficient parallelization for 
these systems, which may have additional challenges associated with
their non-shared memory model.
%\footnote{Non-shared memory configurations include CPU blade servers, multi-GPU servers, and clusters}.
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.  
% WORKING HERE

@@

\subsection{GPU Hardware Concurrency} \vspace{-0.05in}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
For our GPU implementations, we target NVIDIA Fermi GPU architecture (\eg, the GTX 480
GPU used in this paper).  An explanation of the GPU programming model (CUDA) at a useful level of completeness would take more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. 
%\footnote{A discussion of the CUDA programming model can be found in the NVIDIA C for CUDA Programming Guide}
CUDA programmable GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

% GPU CACHE DISCUSSION
% for L2 cache
% 768KB = 786432 bytes
% 64*64*4 = 2**14 bytes = 16384 bytes per image
% 48 images
Each SMP contains its own L1-cache, which is divided between hardware managed
and software managed ("shared") memory.  Additionally, all SMPs share a common
L2-cache.  For our system, each SMP has 64 KiB of L1 cache, and all SMP's share
768KB of L2 cache. 
%The cache hierarchy 
The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is balanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.  A scale drawing of the caches available on the GPU can be seen in Figure \ref{fig:gpu_caches}.
The GPU has its own memory system, and any data the GPU uses must first be
transferred from CPU DRAM to GPU DRAM over PCI-Express.  For our application,
this transfer overhead can be amortized over a large amount of computation and is
not a major concern.

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced concurrency.\footnote{
In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em thread-level} concurrency.  Note that the GPU provides more
concurrency than a CPU at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

@@

Given the hardware concurrency in modern parallel computing architectures,
next, we shall study the optimal mappings of the algorithm parallelism in face
recognition to the multi-core CPUs and GPUs, respectively.  In this section, we
first focus on the face alignment stage (see Figure \ref{fig:pipeline}).

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation
$\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose
variation w.r.t. each training class $A_i$, which leads to the minimal sparsity
in error $\ee$ after the alignment. Note that directly solving
\eqref{eq:l1min_alignment} is inefficient since it is a non-convex problem
and may exhibit local minima.
However, given a good initial guess of the
transformation $\tau$ (\eg, provided by an accurate face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
\small
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

\subsection{GPU Implementation} \vspace{-0.05in}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, cache is relatively unimportant on the GPU for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes is its ability to sustain a very
high bandwidth to DRAM.  This bandwidth is achieved by having a very large number of
threads issuing interleaved memory requests.  Solving many alignment problem instances
concurrently increases the number of threads that can be used.

Therefore, our recommended GPU implementation is strikingly similar in spirit
to our recommended CPU implementation: on the CPU we use the cores to solve
multiple instances concurrently, and on the GPU we use multiple SMP's for the 
same purpose.\footnote{In CUDA terms, our proposed alignment stage
implementation consists of a single kernel that performs alignment for all
subject classes.  Each instance of the alignment problem is assigned its own
thread block, and the GPU hardware schedules as many thread blocks to run
concurrently as the hardware will allow.} The number of subject classes that
are actually scheduled to run concurrently is chosen by the GPU hardware, 
but can be indirectly influenced by manual tuning of the code (i.e., the kernel launch
configuration in CUDA terms). We have empirically determined
that performance is highest with 5-7 subject classes aligned simultaneously on each SMP.
%This mapping of the problem-level parallelism onto the SMP-level
%concurrency is illustrated in Figure \ref{fig:alignment_mapping_gpu}.
%\begin{figure}
%\centering
%\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
%\caption{Proposed mapping of alignment parallelism onto GPU concurrency}
%\label{fig:alignment_mapping_gpu}
%\end{figure}

Several other aspects of the implementation merit discussion.  
First of all, we take advantage of the GPU's special hardware dedicated to bilinear interpolation
for the computation of $b(\tau)$ and $J(\tau)$, which essentially consist of resampled
versions of the test image and its derivatives.
Second, since there are no standard CUDA libraries that work at the SMP level, we use a
custom routine for computing $B_j^\dagger = (B_j^TB_j)^{-1} B^T = G^{-1} B_j^T$, with $G$
inverted using Gaussian elimination with partial pivoting.  To achieve
precision comparable to the single precision LAPACK routines in Intel's MKL
with this simplistic algorithm, we use double precision.  Since $G$ is only $n
\times n$ and $B_j^\dagger$ is only computed once per optimization problem, the
cost of the inversion is dwarfed by the cost of other steps.  
Similarly, sums and dot-products of large vectors are also
performed in double precision.
%The lack of a atomic floating point addition on the GPU
%significantly reduces the perfomance of this step of the algorithm.
%In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
%implementations running on synthetic and real image data.

\section{Parallelization of the Face Recognition Stage} \vspace{-0.05in}
\label{sec:recognition}
After the face alignment stage is complete, the 20 subject classes with lowest
alignment error are selected for recognition, $\bb$ and $A_i$ are re-sampled to a common alignment
using $\tau_i$, and $A_i$ are concatenated into a new matrix $A$.
A sparse representation of $\bb$ w.r.t $A$ is then recovered
by solving a single $\ell_1$-min problem, as shown
in \eqref{eq:l1min_denoise}, using Algorithm \ref{alg:alm_rec}.  The
coefficients $\x$ are then used to compute error residuals that are used for
classification.  
%This recognition algorithm has
%been extensively studied in
%\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}.
In this stage of the algorithm, there is no problem-level
parallelism to exploit, so this section will discuss how to exploit pixel-level parallelism
on both CPU and GPU hardware.
Then in Section \ref{sec:experiment}, we will benchmark the
performance of the two architectures and further demonstrate the speed
gains achieved by our proposed implementations over previously published implementations.

@@

First, in order to measure the impact of solving many
$\ell_1$ problems-per concurrently on the GPU, we benchmark three implementations of the
alignment $\ell_1$-min solver with $A$ of size $5120 \times 32$, and a fixed
$50$ inner loop iterations for each of $50$ outer loop iterations. The runtime
on a GTX480 GPU is averaged over a large number of trials, which are run
sequentially or concurrently depending on the implementation.
The results are shown in Table \ref{tbl:ubench}.
Our proposed parallelization of the $\ell_1$-min used in the alignment
stage is {\em eight times} faster than an implementation solving a single
problem at a time using the stock BLAS libraries.
\footnote{The streams implementation is limited significantly by a 
cap on the number of concurrent streams in the current version of CUDA}
%The results are shown below:
\begin{table}[t!]
\caption{\small Benchmark of implementations of alignment $\ell_1$-min.}
\small
\begin{tabular}{|l|c|}
\hline
Sequential solver using CUBLAS & 302\,ms \\
\hline
One problem per SMP using CUDA streams & 70\,ms  \\
\hline
Four problems per SMP using single kernel & {\bf 36\,ms} \\
\hline
\end{tabular} \vspace{-0.1in}
\label{tbl:ubench}
\end{table}

@@

Figure \ref{fig:alignment_stage_runtime} shows the average run time of the CPU and GPU implementations to align
a query image against all the subject classes separately. We vary the total number of
subject classes to show how the algorithms scale. The plateaus seen in the GPU curve
result from the GPU hardware scheduling the computation of alignment
problems in concurrent batches, but the overall trend is linear as expected.
The manually threaded CPU implementation slightly outperforms the GPU implementation,
and it beats the \emph{naive}
library threaded implementation by a wide margin.
The new implementation can align the query
image in $\approx 40$ ms per subject, 
while the fastest previously published result \cite{WagnerA2011-PAMI}
required $\approx 600$ ms seconds per subject in a similar setting.

@@

\section{Conclusion}\vspace{-0.05in}
We have demonstrated that on both CPU and GPU algorithms, parallelizations of
ALM that solve multiple face alignment problems concurrently are significantly
faster than implementations that rely purely on vendor-supplied BLAS libraries.
Furthermore, thanks to a combination of faster hardware, and more efficient use
of the hardware, we have demonstrated dramatic improvements in recognition
speed over previously reported implementations.
As CPU manufacturers increase the number of cores and vector widths, and as GPU 
manufacturers increase the amount
of cache, both architectures are rapidly converging towards an architectural balance
that increasingly favorable for $\ell_1$-min based face recognition.
% of the same pipeline.  Whereas
%earlier implementations required over a second to align a test image against
%just one of the 250 subjects in Multi-PIE session 1, the proposed 
%implementations can perform the same task in 
%{\bf under 10 seconds}!.  
%The parallelization techniques presented here make it feasible to implement an
%access control system for hundreds of subjects on a handful of purpose-build
%multi-GPU workstations.

%We argue that in addition to vendor supplied libraries
%that attempt to leverage both levels of concurrency of the GPU, there is a need
%for standard libraries that operate at the SMP level of concurrency.

>>>>

Among its many applications, $\ell_1$-min has been recently used to reformulate
image-based face recognition as a sparse representation problem
\cite{WrightJ2009-PAMI}.  If we stack the $d$ pixels of the training images of $K$ subject
classes into the columns of matrices $(A_1\in\Re^{d\times n_1}, \cdots, A_K\in\Re^{d\times n_K})$, combine
the matrices into a larger matrix $A = [A_1, \cdots, A_K]\in\Re^{d\times n}$, and arrange the pixels of a new
query image into a vector $\bb\in\Re^d$, \emph{sparsity-based
classification} (SBC) solves the following minimization problem:
\begin{equation}
\min_{\xx, \ee} \| \xx \|_1 + \|\ee\|_1 \quad \subj \quad \bb = A \xx + \ee.
\label{eq:l1min_denoise}
\end{equation}
If the sparsest solutions for $\x$ and $\e$ are recovered, $\ee$ provides a
means to compensate for pixels that are corrupted due to occlusion of some part of the query
image, and the dominant nonzero coefficients of $\xx$ reveal the membership of
$\bb$ based on the training image labels associated with $A$. 
%Since $A$ contains many images per user taken under different illuminations, 
%$A_i\x_i$ acts as a linear illumination model for the test image $\bb$.

@@

Finally, we present extensive benchmarks to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of subjects or more will likely require computation to be
distributed over a cluster or network of computers. These systems are not addressed here.
The source code of our implementations, which target NVIDIA Fermi GPUs
as well as Intel x86-64 CPUs, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.\vspace{-0.1in}}
\label{fig:pipeline}
\end{figure}

@@

Our discussion and experiments will address the most common hardware
configuration for engineering workstations: a motherboard with two quad-core
processors and a PCIe card with a single high-end GPU.  
Recognition involving more than a few hundred subjects with contemporary hardware
would require a server (or cluster) configuration with an expandable 
number of processors.  We will not address efficient parallelization for 
these systems, which may have additional challenges associated with
their non-shared memory model.
%\footnote{Non-shared memory configurations include CPU blade servers, multi-GPU servers, and clusters}.
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.  
% WORKING HERE

@@

\subsection{GPU Hardware Concurrency} \vspace{-0.05in}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
For our GPU implementations, we target NVIDIA Fermi GPU architecture (\eg, the GTX 480
GPU used in this paper).  An explanation of the GPU programming model (CUDA) at a useful level of completeness would take more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. 
%\footnote{A discussion of the CUDA programming model can be found in the NVIDIA C for CUDA Programming Guide}
CUDA programmable GPUs are comprised of several
\emph{streaming multiprocessors} (SM), each of which is roughly analogous to a CPU core.
For Fermi architecture GPUs there are up to 16 SMs, and each SM is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

% GPU CACHE DISCUSSION
% for L2 cache
% 768KB = 786432 bytes
% 64*64*4 = 2**14 bytes = 16384 bytes per image
% 48 images
Each SM contains its own L1-cache, which is divided between hardware managed
and software managed (``shared") memory.  Additionally, all SMs share a common
L2-cache.  For our system, each SM has 64 KiB of L1 cache, and all SMs share
768KB of L2 cache. 
%The cache hierarchy 
The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is balanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.  A scale drawing of the caches available on the GPU can be seen in Figure \ref{fig:gpu_caches}.
The GPU has its own memory system, and any data the GPU uses must first be
transferred from CPU DRAM to GPU DRAM over PCIe.  For our application,
this transfer overhead can be amortized over a large amount of computation and is
not a major concern.

While the programming model for
the SIMD units of each SM is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced concurrency.\footnote{
In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SM-level}
concurrency and {\em thread-level} concurrency.  Note that the GPU provides more
concurrency than a CPU at both levels (14 SMs vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

@@

We will next discuss the parallelism available in the face recognition
pipeline, and propose techniques for exploiting this parallelism on the GPU and CPU hardware 
described in the previous section.  In this section we first focus on the
face alignment stage (see Figure \ref{fig:pipeline}).

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation
$\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose
variation w.r.t. each training class $A_i$, which leads to the minimal sparsity
in error $\ee$ after the alignment. Note that directly solving
\eqref{eq:l1min_alignment} is inefficient since it is a non-convex problem
and may exhibit local minima.
However, given a good initial estimate of the
transformation $\tau$ (\eg, provided by an accurate face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
\small
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

\subsection{GPU Implementation} \vspace{-0.05in}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, cache is relatively unimportant on the GPU for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:gpu_caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes is its ability to sustain a very
high bandwidth to DRAM.  This bandwidth is achieved by having a very large number of
threads issuing interleaved memory requests.  Solving many alignment problem instances
concurrently increases the number of threads that can be used.

Therefore, our recommended GPU implementation is strikingly similar in spirit
to our recommended CPU implementation: on the CPU we use the cores to solve
multiple instances concurrently, and on the GPU we use multiple SMs for the 
same purpose.\footnote{In CUDA terms, our proposed alignment stage
implementation consists of a single kernel that performs alignment for all
subject classes.  Each instance of the alignment problem is assigned its own
thread block, and the GPU hardware schedules as many thread blocks to run
concurrently as the hardware will allow.} The number of subject classes that
are actually scheduled to run concurrently is chosen by the GPU hardware, 
but can be indirectly influenced by manual tuning of the code (i.e., the kernel launch
configuration in CUDA terms). We have empirically determined
that performance is highest with 5-7 subject classes aligned simultaneously on each SM.
%This mapping of the problem-level parallelism onto the SM-level
%concurrency is illustrated in Figure \ref{fig:alignment_mapping_gpu}.
%\begin{figure}
%\centering
%\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
%\caption{Proposed mapping of alignment parallelism onto GPU concurrency}
%\label{fig:alignment_mapping_gpu}
%\end{figure}

Several other aspects of the implementation merit discussion.  
First of all, we take advantage of the GPU's special hardware dedicated to bilinear interpolation
for the computation of $b(\tau)$ and $J(\tau)$, which essentially consist of resampled
versions of the test image and its derivatives.
Second, since there are no standard CUDA libraries that work at the SM level, we use a
custom routine for computing $B_j^\dagger = (B_j^TB_j)^{-1} B^T = G^{-1} B_j^T$, with $G$
inverted using Gaussian elimination with partial pivoting.  To achieve
precision comparable to the single precision LAPACK routines in Intel's MKL
with this simplistic algorithm, we use double precision.  Since $G$ is only $n
\times n$ and $B_j^\dagger$ is only computed once per optimization problem, the
cost of the inversion is dwarfed by the cost of other steps.  
Similarly, sums and dot-products of large vectors are also
performed in double precision.
%The lack of a atomic floating point addition on the GPU
%significantly reduces the perfomance of this step of the algorithm.
%In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
%implementations running on synthetic and real image data.

\section{Parallelization of the Face Recognition Stage} \vspace{-0.05in}
\label{sec:recognition}
After the face alignment stage is complete, the 20 subject classes with lowest
alignment error are selected for recognition, $\bb$ and $A_i$ are re-sampled to a common alignment
using $\tau_i$, and $A_i$ are concatenated into a new matrix $A$.
A sparse representation of $\bb$ w.r.t $A$ is then recovered
by solving a single $\ell_1$-min problem, as shown
in \eqref{eq:l1min_denoise}, using Algorithm \ref{alg:alm_rec}.  The
coefficients in $\x$ are then used to compute error residuals that are used for
classification.  
%This recognition algorithm has
%been extensively studied in
%\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}.
In this stage of the algorithm, there is no problem-level
parallelism to exploit, so this section will discuss how to exploit pixel-level parallelism
on both CPU and GPU hardware.
Then in Section \ref{sec:experiment}, we will benchmark the
performance of the two architectures and further demonstrate the speed
gains achieved by our proposed implementations over previously published implementations.

@@

First, in order to measure the impact of solving many
$\ell_1$ problems-per concurrently on the GPU, we benchmark three implementations of the
alignment $\ell_1$-min solver with $A$ of size $5120 \times 32$, and a fixed
$50$ inner loop iterations for each of $50$ outer loop iterations. The runtime
on a GTX480 GPU is averaged over a large number of trials, which are run
sequentially or concurrently depending on the implementation.
The results are shown in Table \ref{tbl:ubench}.
Our proposed parallelization of the $\ell_1$-min used in the alignment
stage is {\em eight times} faster than an implementation solving a single
problem at a time using the stock BLAS libraries.
\footnote{The streams implementation is limited significantly by a 
cap on the number of concurrent streams in the current version of CUDA}
%The results are shown below:
\begin{table}[t!]
\caption{\small Benchmark of implementations of alignment $\ell_1$-min.}
\small
\begin{tabular}{|l|c|}
\hline
Sequential solver using CUBLAS & 302\,ms \\
\hline
One problem per SM using CUDA streams & 70\,ms  \\
\hline
Four problems per SM using single kernel & {\bf 36\,ms} \\
\hline
\end{tabular} \vspace{-0.1in}
\label{tbl:ubench}
\end{table}

@@

Figure \ref{fig:alignment_stage_runtime} shows the average run time of the CPU and GPU implementations to align
a query image against all the subject classes separately. We vary the total number of
subject classes to show how the algorithms scale. The plateaus seen in the GPU curve
result from the GPU hardware scheduling the computation of alignment
problems in concurrent batches, but the overall trend is linear as expected.
The manually threaded CPU implementation slightly outperforms the GPU implementation,
and it surpasses the \emph{naive}
library threaded implementation by a wide margin.
The new implementation can align the query
image in $\approx 40$ ms per subject, 
while the fastest previously published result \cite{WagnerA2011-PAMI}
required $\approx 600$ ms seconds per subject in a similar setting.

@@

\section{Conclusion}\vspace{-0.05in}
We have demonstrated that on both CPU and GPU algorithms, parallelizations of
ALM that solve multiple face alignment problems concurrently are significantly
faster than implementations that rely purely on vendor-supplied BLAS libraries.
Furthermore, thanks to a combination of faster hardware and more efficient use
of the hardware, we have demonstrated dramatic improvements in recognition
speed over previously reported implementations.
As CPU manufacturers increase the number of cores and vector widths, and as GPU 
manufacturers increase the amount
of cache, both architectures are rapidly converging towards an architectural balance
that increasingly favorable for $\ell_1$-min based face recognition.
% of the same pipeline.  Whereas
%earlier implementations required over a second to align a test image against
%just one of the 250 subjects in Multi-PIE session 1, the proposed 
%implementations can perform the same task in 
%{\bf under 10 seconds}!.  
%The parallelization techniques presented here make it feasible to implement an
%access control system for hundreds of subjects on a handful of purpose-build
%multi-GPU workstations.

%We argue that in addition to vendor supplied libraries
%that attempt to leverage both levels of concurrency of the GPU, there is a need
%for standard libraries that operate at the SM level of concurrency.
